### 2013-11-06 docs/fsml/paper.tex
Added the DSL primer, first version, to the repo.

\documentclass[preprint,authoryear,12pt]{noelsarticle}
%\documentclass[preprint,12pt]{elsarticle}
%\documentclass[final,3p,times,authoryear]{elsarticle}
%\documentclass[final,3p,times]{elsarticle}

\usepackage{tikz}
\usepackage{hyperref}
\usepackage{verbatim}
\usepackage{comment}
\usepackage{boxedminipage}
\usepackage{amssymb}
\usepackage{listings,lsthaskell,lstprolog,lstegl,lstesl,lstfsml}

\newcommand{\m}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\concept}[1]{\emph{#1}}
\definecolor{codebackground}{rgb}{0.97,0.97,0.97}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{keyword}{rgb}{0.5,0.0,0.5}
\definecolor{comment}{rgb}{0.3,0.5,0.3}
\definecolor{code}{rgb}{0.0,0.0,0.3}
\definecolor{ncode}{rgb}{0.0,0.3,0.3}
\definecolor{scode}{rgb}{0.3,0.0,0.3}
\definecolor{static}{rgb}{0.0,0.0,0.3}
\newcommand{\ncode}[1]{\texttt{\textup{\color{ncode}#1}}}
\newcommand{\jcode}[1]{\texttt{\textup{\color{code}#1}}}
\newcommand{\scode}[1]{\texttt{\textup{\color{scode}#1}}}
\newcommand{\static}[1]{\texttt{\textsl{\color{static}#1}}}

\newcommand{\codefigure}[3]{
\begin{figure}[t!]
\begin{boxedminipage}{\hsize}
\mbox{}\hfill{}{\small\textsf{\href{http://github.com/slebok/slepro/tree/master/#2}{#2}}}
\lstinputlisting[language=#3]{../../#2}
\end{boxedminipage}
\caption{#1.}
\label{#2}
\end{figure}}

\lstset{%
  fontadjust=true,%
  showstringspaces=false,%
  keywordstyle=\color{keyword},%
  commentstyle=\color{comment}\rmfamily\itshape\small,%
  columns=fullflexible,%
  tabsize=2,
%  numbers=left,                   % where to put the line-numbers
  numberstyle=\tiny\color{gray},  % the style that is used for the line-numbers
  backgroundcolor=\color{codebackground},
  stepnumber=2,                   % the step between two line-numbers
%  numbersep=5pt,                  % how far the line-numbers are from the code
 % aboveskip=1.5\medskipamount,%
 % belowskip=1.5\medskipamount,
aboveskip = 0pt,
belowskip = 0pt,
  %breaklines=true
}

\lstdefinestyle{haskell}{
  language=Haskell,
  basicstyle=\small\color{ncode}\ttfamily,%
}

\lstdefinestyle{prolog}{
  language=Prolog,
  basicstyle=\small\color{ncode}\ttfamily,%
}

\newcommand{\wikipediafn}[2]{\footnote{\url{#1}---last visited #2.}}
\newcommand{\ooo}[1]{\textsf{101#1}}

%% \biboptions{comma,round}
% \biboptions{}

\begin{document}

\begin{frontmatter}

\title{Another DSL primer}

\author{Ralf L\"ammel}

\address{Software Languages Team\\University of Koblenz-Landau, Germany}

\begin{abstract}
  We model a domain-specific language FSML for Finite State Machines
  (FSMs). The language model encompasses concrete a textual syntax, a
  term-based abstract syntax, a graph-based visual syntax, extra
  constraints for well-formedness, a simulation semantics, and a
  code-generation semantics for representing and executing FSMs in
  Java. The key motivation for the present model of FSML to be
  explainable exhaustively in terms of simple formalisms and idioms,
  all layered on top of Prolog.

\bigskip

\noindent
\textbf{Acknowledgement}: {\small This document and the underlying
  development are parts of a broader effort on language modeling and
  software language engineering---joint work with \emph{Anya Helene
    Bagge}, University of Bergen. Helpful interaction with and
  feedback by Andrei Varanovich, University of Koblenz-Landau, is also
  gratefully acknowledged.}

\medskip

\noindent
\textbf{Version}

0.000001 as of 4 November 2013.

\medskip

\noindent
\textbf{Repository location of document}:

\url{https://github.com/slebok/slepro/tree/master/docs/fsml}.
\end{abstract}

% \begin{keyword} 
% ???
% \end{keyword}

\end{frontmatter}

\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\tableofcontents

\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introducing FSML}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[t!]
\vspace{-77\in}
\includegraphics[width=\textwidth]{../../languages/fsml/sample-turnstile.pdf}
\vspace{-150\in}
\caption{A finite state machine for turnstiles.}
\label{F:turnstile}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Consider the visual representation of a finite state machine (FSM) for
a turnstile (as used in subway systems). The idea is, of course, that
the customer is supposed to insert a ticket before passing the
turnstile. Thus, the FSM assumes the initial state `locked'. (The
initial state is highlighted by using a filled ellipse.) The input
`ticket' (meaning the insertion of a ticket) causes a transition to
the state `unlocked' and the action `collect'. In this new state, the
input `pass' (meaning the attempt to pass the turnstile) causes a
transition back to the state `locked'. If the input `pass' was made in
the state `locked', then this causes a transition to the state
`exception' and an action `alarm'. The state `exception' rejects the
input `ticket'. The input `pass' does, of course, keeps one in the
state `exception'. The `alarm' can be muted with the input `mute'. We
return to the normal `locked' state upon the input `release'.

We will model a language FSML (FSM language) for modeling
FSMs. Specifically, we will model several aspects of FSML:

\begin{itemize}
\item The visual syntax as used in the figure; see \S\ref{S:visual}.
\item A textual syntax as an alternative concrete syntax; see \S\ref{S:textual}.
\item An abstract syntax useful for representation in programs; see \S\ref{S:abstract}.
\item Extra well-formedness constraints on the abstract syntax; see \S\ref{S:ok}.
\item A reference semantics for the simulation of FSMs; see \S\ref{S:semantics}.
\item A code generator translating FSMs into OO programs; see \S\ref{S:generator}.
\end{itemize}

This primer is concluded by some `food for thought' in
\S\ref{S:concl}. That is, we state shortcomings or issues of
incompleteness of the language model approach at hand, thereby also
motivating reproductions of the model as well as variations on the
model, possibly leveraging diverse metaprogramming settings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The concrete textual syntax of FSML}
\label{S:textual}

\codefigure{%
The FSM of \autoref{F:turnstile} in concrete textual syntax}{%
languages/fsml/sample-turnstile.fsml}{%
fsml}

A textual syntax may break down an FSM into `state declarations',
i.e., states with the associated transitions. Each transition
identifies the relevant input, the optional action, and the target
state. In fact, if the target state is omitted, then this is taken to
mean that the transition's target is the current state. See
\autoref{languages/fsml/sample-turnstile.fsml} for an illustration.

\codefigure{%
Context-free grammar for the concrete syntax of FSML}{%
languages/fsml/cs.egl}{%
egl}

Let us define the syntax in terms of a context-free grammar. We use an
EGL grammar, where EGL stands for extended grammar language and can be
regarded as a variation on the Extended Backus Naur Form. EGL is an
ad-hoc grammar notation that is provided by the same project that
hosts the present FSML model. In particular, EGL grammars can be
executed for the purpose of parsing, but we omit all such details
here. See \autoref{languages/fsml/cs.egl} for the grammar. It can be
assumed that the concrete textual syntax, as illustrated in
\autoref{languages/fsml/sample-turnstile.fsml}, can be parsed indeed
with the present grammar.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The abstract syntax of FSML}
\label{S:abstract}

Eventually, we want to manipulate FSMs in (Prolog) programs. To this
end, we need a suitable abstract syntax. As an aside, there exist
metaprogramming systems that can leverage the concrete syntax
directly, without any need for a separate abstract syntax, but we will
adopt the simpler approach here to indeed use an abstract syntax. In
fact, we assume that abstract syntax trees are (Prolog) terms as in
the sense of term algebras or algebraic specifications.

\codefigure{%
The FSM of \autoref{F:turnstile} in abstract syntax}{%
languages/fsml/sample-turnstile-fsml.term}{%
prolog}

See \autoref{languages/fsml/sample-turnstile-fsml.term} for an
illustration. The abstract syntax describes a FSM as a list of state
declarations, which are in turn tuples (triplets) of the form (\m{V},
\m{Id}, \m{Ts}) where \m{V} is a Boolean saying whether the state is
initial, \m{Id} is the assigned id (name) of the state, and \m{Ts} is
a list of transitions. Each transition is a triplet of the form
(\m{In}, \m{A}, \m{To}) where \m{In} is the input for the transition,
\m{A} is the optional action and \m{To} is the target state. The
optionality of the action is represented such that the missing action
is represented by `[]' and the present action, e.g., `eject', is
represented by `[eject]'.

\codefigure{%
Signature for the abstract syntax of FSML}{%
languages/fsml/as.esl}{%
esl}

Let us define the abstract syntax in terms of a signature that defines
all valid FSM terms. We use an ESL signature, where ESL stands for
extended signature language and can be regarded as (quite) a variation
on notations familiar from algebraic specification or (less) a
variation on type systems used in logic or functional programming. ESL
is an ad-hoc signature notation that is provided by the same project
that hosts the present FSML model. In particular, ESL grammars can be
executed for conformance checking to see whether a given term conforms
to a given signature, but we omit all such details here. See
\autoref{languages/fsml/as.esl} for the signature. The term in
\autoref{languages/fsml/sample-turnstile-fsml.term} conforms indeed
to the present signature.

\codefigure{%
Concrete to abstract syntax mapping for FSML}{%
languages/fsml/cs-to-as.pro}{%
prolog}

As we intend to parse FSMs via the concrete textual syntax and to
manipulate FSMs though via the abstract term-based syntax, we need a
mapping from the former to the latter. Of course, both syntaxes are
very close to each other in terms of the nonterminals or types defined
and in terms of the structural breakdown. However, some fine details
need to be declared explicitly via a mapping so that parse trees
according to the earlier grammar are precisely mapped to terms that
conform to the signature for the abstract syntax. The mapping is
defined by a Prolog predicate \emph{fsmlMapping}; see the Prolog
clauses in \autoref{languages/fsml/cs-to-as.pro}.

The syntax definition approach of the present project assumes such
mappings are indeed predicates with three arguments \m{N}, \m{T1},
\m{T2} as follows. \m{N} is the nonterminal (according to the grammar
for the concrete syntax) whose parse trees are to be mapped, i.e.,
rewritten. \m{T1} is the (imploded) parse tree that just
systematically follows the structure of the grammar. \m{T2} is the
term that should be used in place of \m{T1}. The starting point is an
identity mapping; the mapping predicate only overwrites those cases
where the identity mapping is not appropriate. In
\autoref{languages/fsml/cs-to-as.pro}, the first clause replaces the
list-of-chars representation of FSM names by proper atoms (`strings').
The second clause replaces `initial' by `true' (and `noninitial' by
`false'); it also fills in the target state, when it is missing,
thereby taking care of some syntactic sugar, i.e., the permitted
omission of the target state when it equals the source state of a
transition.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Constraints on the abstract syntax}
\label{S:ok}

The concrete and abstract syntaxes so far do not cover some of the
constraints that we would naturally apply to FSMs. For instance, we
did not rule out so far that FSMs have multiple initial states, which
is probably not useful. Context-free grammars and signatures are
notoriously limited in expressiveness to model such constraints,
whereas they are convenient for the basic definition of structure. We
need to add extra constraints on top of (say) the abstract syntax
definition. In some communities, it is common to actually consider
these constraints to form part of the abstract syntax whereas we
follow another common view that such constraints are considered
separately as in well-formedness or well-typedness judgements used in
type systems for programming languages.

\codefigure{%
Overview of constraints imposed on FSML's abstract syntax}{%
languages/fsml/ok.pro}{%
prolog}

See \autoref{languages/fsml/ok.pro} for the Prolog specification
listing the named constraints that we consider here. Each constraint
gives rise to a separate predicate considered below. Here is short
summary. Constraint \emph{fsmSingleInitial} is meant to ensure that
there is only a single state declaration for an initial
state. Constraint \emph{fsmDistinctIds} is meant to ensure that the
state ids of the state declarations are distinct. Constraint
\emph{fsmResolvable} is meant to ensure that all referenced target
states are declared. Constraint \emph{fsmDeterministic} is meant
to ensure that each given input uniquely defines the target state to
be transitioned to, if any. Constraint \emph{fsmReachable} is meant to
ensure that all declared states are reachable from the
initial state. In \autoref{languages/fsml/ok.pro}, all constraint
predicates are surround by an application of the meta-predicate
\emph{require/1}, which is simply there for better error
reporting. That is, if the argument of \emph{require/1} fails, then
this is reported immediately, as opposed to simply propagating failure
silently.

\codefigure{%
There is only a single state declaration for an initial state}{%
languages/fsml/ok/initial.pro}{%
prolog}

\codefigure{%
The state ids of the state declarations are distinct}{%
languages/fsml/ok/distinct.pro}{%
prolog}

\codefigure{%
All referenced target states are declared}{%
languages/fsml/ok/resolvable.pro}{%
prolog}

\codefigure{%
Each given input uniquely defines the target state to be transitioned to, if any}{%
languages/fsml/ok/deterministic.pro}{%
prolog}

\codefigure{%
All declared states are reachable from the initial state}{%
languages/fsml/ok/reachable.pro}{%
prolog}

\codefigure{%
Fixed point computation needed in \autoref{languages/fsml/ok/reachable.pro}}{%
languages/fsml/ok/closure.pro}{%
prolog}

All constraints are specified in
\autoref{languages/fsml/ok/initial.pro}--\autoref{languages/fsml/ok/closure.pro}. The
constraints are relatively simple to express using Prolog's
meta-predicate \emph{findall/3} which serves here for queries into the
terms of the abstract syntax. For instance, in
\autoref{languages/fsml/ok/initial.pro}, we find all state ids from
state declarations with `true' for the initial component. Once this
set is retrieved that the resulting list is of length 1; thus, there
is exactly one initial.

The constraint for reachability
(\autoref{languages/fsml/ok/reachable.pro}--\autoref{languages/fsml/ok/closure.pro})
is somewhat interesting, as we have to model a fixed point computation
to find all states by repeated consideration of transition as to
whether the set of known to be reachable states implies more reachable
states. The fixed point criterion is here that we do not find
additional states; see \autoref{languages/fsml/ok/closure.pro}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The reference semantics for FSML}
\label{S:semantics}

TBD

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The code generator for FSML}
\label{S:generator}

TBD

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The visual syntax of FSML}
\label{S:visual}

TBD

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Food for thought}
\label{S:concl}

TBD

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{\small

%\nocite{FreeDictionary}
%\setlength{\bibsep}{4pt}
%\bibliographystyle{elsarticle-harv}
%\bibliographystyle{elsarticle-num}
%\bibliography{paper}

}

\end{document}

### 2013-11-09 docs/fsml/paper.tex
Completed the DSL primer, first draft.

@@
 \usepackage{amssymb}
-\usepackage{listings,lsthaskell,lstprolog,lstegl,lstesl,lstfsml}
+\usepackage{listings,lsthaskell,lstprolog,lstegl,lstesl,lstfsml,lstjava,lstdot}
 
 \newcommand{\m}[1]{\ensuremath{\mathit{#1}}}
+\newcommand{\slepro}{\textsc{SlePro}}
 \newcommand{\concept}[1]{\emph{#1}}
@@
 \begin{boxedminipage}{\hsize}
-\mbox{}\hfill{}{\small\textsf{\href{http://github.com/slebok/slepro/tree/master/#2}{#2}}}
+\mbox{}\hfill{}{\small\textit{\href{http://github.com/slebok/slepro/tree/master/#2}{#2}}}
 \lstinputlisting[language=#3]{../../#2}
@@
 \label{#2}
+\medskip
 \end{figure}}
@@
   showstringspaces=false,%
+  basicstyle=\small\sffamily,%
   keywordstyle=\color{keyword},%
@@
   %breaklines=true
-}
-
-\lstdefinestyle{haskell}{
-  language=Haskell,
-  basicstyle=\small\color{ncode}\ttfamily,%
-}
-
-\lstdefinestyle{prolog}{
-  language=Prolog,
-  basicstyle=\small\color{ncode}\ttfamily,%
 }
@@
   explainable exhaustively in terms of simple formalisms and idioms,
-  all layered on top of Prolog.
+  all layered on top of Prolog. The FSML development is part of the
+  \slepro{} project; see \url{https://github.com/slebok/slepro}.
 
@@
 
-0.000001 as of 4 November 2013.
+0.00001 as of 8 November 2013.
 
@@
 \end{abstract}
-
-% \begin{keyword} 
-% ???
-% \end{keyword}
 
@@
 This primer is concluded by some `food for thought' in
-\S\ref{S:concl}. That is, we state shortcomings or issues of
-incompleteness of the language model approach at hand, thereby also
-motivating reproductions of the model as well as variations on the
-model, possibly leveraging diverse metaprogramming settings.
+\S\ref{S:concl}. That is, we critically discuss the language model at
+hand as well as the underlying approach, thereby also motivating
+reproductions of the model as well as variations on the model,
+possibly leveraging diverse technologies and techniques.
+
+The FSML development is part of the \slepro{} project. All software
+artifacts of this document and all support infrastructure is available
+through the repository of the project:
+
+\begin{center}
+\url{https://github.com/slebok/slepro}
+\end{center}
 
@@
 Let us define the syntax in terms of a context-free grammar. We use an
-EGL grammar, where EGL stands for extended grammar language and can be
-regarded as a variation on the Extended Backus Naur Form. EGL is an
-ad-hoc grammar notation that is provided by the same project that
-hosts the present FSML model. In particular, EGL grammars can be
-executed for the purpose of parsing, but we omit all such details
+EGL grammar, where EGL stands for `extended grammar language' and can
+be regarded as a variation on the Extended Backus Naur Form. EGL is an
+ad-hoc grammar notation that is provided by the \slepro{} project; see
+\S\ref{A:egl} for a self-description. In particular, EGL grammars can
+be executed for the purpose of parsing, but we omit all such details
 here. See \autoref{languages/fsml/cs.egl} for the grammar. It can be
@@
 all valid FSM terms. We use an ESL signature, where ESL stands for
-extended signature language and can be regarded as (quite) a variation
-on notations familiar from algebraic specification or (less) a
-variation on type systems used in logic or functional programming. ESL
-is an ad-hoc signature notation that is provided by the same project
-that hosts the present FSML model. In particular, ESL grammars can be
-executed for conformance checking to see whether a given term conforms
-to a given signature, but we omit all such details here. See
+`extended signature language' and can be regarded as (quite) a
+variation on notations familiar from algebraic specification or (less)
+a variation on type systems used in logic or functional
+programming. ESL is an ad-hoc signature notation that is provided by
+the \slepro{} project; see \S\ref{A:esl} for a syntax definition of
+the signature notation. In particular, ESL grammars can be executed
+for conformance checking to see whether a given term conforms to a
+given signature, but we omit all such details here. See
 \autoref{languages/fsml/as.esl} for the signature. The term in
-\autoref{languages/fsml/sample-turnstile-fsml.term} conforms indeed
-to the present signature.
+\autoref{languages/fsml/sample-turnstile-fsml.term} conforms indeed to
+the present signature.
 
@@
 
-The syntax definition approach of the present project assumes such
+The syntax definition approach of the \slepro{} project assumes such
 mappings are indeed predicates with three arguments \m{N}, \m{T1},
@@
 
-TBD
+So far we relies on an intuitive understanding of the FSML semantics,
+which is reasonable in so far that the notion of finite state machines
+is pretty fundamental and familiar. Nevertheless, we are bound to
+provide a precise and executable reference semantics of the language
+which can then be hold accountable by other language-based software
+components, e.g., a code generator (see next section) or a
+refactoring.
+
+We speak of a reference semantics here to emphasize that this
+semantics may not be practically useful immediately. That is, we
+will define the semantics of a FSM by a simulation judgement which
+assumes that all input is provided upfront and all output (the action
+sequence) is provided upon completion of input processing. This is a
+batch model, which is useful in a formal semantics, but it is not
+useful for using FSMs in interactive systems.
+
+\codefigure{%
+Semantics of FSML}{%
+languages/fsml/simulation.pro}{%
+prolog}
+
+See \autoref{languages/fsml/simulation.pro} for the semantics; the
+judgements are expressed in Prolog. The clause at the top selects the
+initial state of the FSM and enters the simulation with that state and
+the complete input. The remaining two clauses describe the iteration
+over the input such that a suitable transition is looked up in each
+step and action (if any) and the new state are paired up as an element in
+the output so that simulation proceeds with the new state and the
+remaining input.
 
@@
 
-TBD
+A practically useful interpretation of FSML would be achieved, if we
+enabled interactive execution of the FSMs in the context of some
+programming environment. We pick Java as our target here. We describe
+a code generator which translates FSMs into Java code. The generated
+code can be augmented with hand-written components specifically for
+the intended semantics of actions. The Java representation of FSMs
+meets the requirement that execution is step-by-step. That is, a
+transition is transparently realized including the potential
+performance of an action whenever an input item arrives.
+
+There are many different ways to generate imperative/OO code for
+FSMs. We pick one particular option here--without loss of
+generality. The chosen option can be said to be data-centric in that
+FSMs are represented in an appropriate container-based data structure,
+whereas the execution of FSMs relies on the interpretation of said
+data structure. Further, the programmer-defined meaning of actions is
+provided by a handler which interpreters action names.
+
+\codefigure{%
+Java representation of the states for the turnstile example}{%
+languages/fsml/java/State.java}{%
+java}
+
+\codefigure{%
+Java representation of the inputs for the turnstile example}{%
+languages/fsml/java/Input.java}{%
+java}
+
+\codefigure{%
+Java representation of the actions for the turnstile example}{%
+languages/fsml/java/Action.java}{%
+java}
+
+\codefigure{%
+Code generation for state type}{%
+languages/fsml/to-java/state.pro}{%
+prolog}
+
+\autoref{languages/fsml/java/State.java}--\autoref{languages/fsml/java/Action.java}
+shows (generated) Java enum types for the states, inputs, and
+actions. Obviously, these sets are trivially obtainable from the
+abstract syntactical representation of an FSM. In
+\autoref{languages/fsml/to-java/state.pro}, this is illustrated for
+the enum type for states. The predicate collects all (declared) state
+ids and assembles an enum type, subject to a suitable abstract syntax
+of Java; see \S\ref{A:java} for a summary of the (abstract) Java
+syntax used by the code generator. Finally, the Java declaration is
+pretty printed; see the use of \emph{ppJavaDecl/2} in the figure,
+subject to a pretty printer for the Java language. The generator
+predicates for the enum types for inputs and actions are quite
+similar.
+
+\codefigure{%
+Generic interface for action handers}{%
+languages/fsml/java/Handler.java}{%
+java}
+
+\codefigure{%
+A trivial interpretation of the turnstile actions}{%
+languages/fsml/java/TurnstileHandler.java}{%
+java}
+
+\autoref{languages/fsml/java/Handler.java} shows a generic, i.e.,
+FSM-independent interface for defining handlers for FSM
+actions. \autoref{languages/fsml/java/TurnstileHandler.java} shows a
+trivial implementation of said interface for the turnstile
+example. The actions simply print out the names of the actions, but it
+is clear that arbitrary functionality can be plugged into FSM
+execution in this manner.
+
+\codefigure{%
+Generic class for FSM execution}{%
+languages/fsml/java/Stepper.java}{%
+java}
+
+\autoref{languages/fsml/java/Stepper.java} shows essentially the
+runtime for FSMs. This is generic code which is appropriately
+parameterized in the types for states, inputs, and actions. The class
+relies on a table for holding the transitions in the form of a map
+from states to maps from inputs to pairs of optional actions and
+states. There is an appropriate \emph{add} method which makes it easy
+to insert into this non-basic container. The class is also prepared to
+keep track of and make use of a handler for the involved actions. The
+class is called \emph{Stepper} because its key method is the one for
+performing a step (a transition) based on an actual input item.
+
+\codefigure{%
+Configuration of a stepper for the turnstile example}{%
+languages/fsml/java/TurnstileStepper.java}{%
+java}
+
+\autoref{languages/fsml/java/TurnstileStepper.java} shows the
+generated code for the turnstile-specific \emph{Stepper} subclass. The
+class instantiates the generic parameters of the \emph{Stepper} class
+to the turnstile-specific enum types for states, inputs, and actions,
+which were shown earlier. Also, the class encodes all transitions of
+the turnstile FSM via appropriate calls to the \emph{add} method, as
+part of the constructor. The initial state is also set up and the
+turnstile-specific handler is communicated as well.
+
+\codefigure{%
+Code generation for stepper subclass}{%
+languages/fsml/to-java/stepper.pro}{%
+prolog}
+
+\autoref{languages/fsml/to-java/stepper.pro} shows the code generation
+predicate for stepper subclasses. The predicate refers to more
+elements of the Java language: several expression forms (e.g.,
+`null'), assignment statements, method calls, and class
+declarations. In a final step, the constructed abstract syntactical
+representation is passed to the pretty printer for Java.
 
@@
 
-TBD
+It remains to define the visual syntax of FSML, as it was illustrated
+in \autoref{F:turnstile}. A visual syntax definition may serve these
+major purposes: actual representation (visualization) and
+editing. Indeed, we want this syntax to be complemented by rendering
+support. We do not consider editing here.
+
+\codefigure{%
+The FSM of \autoref{F:turnstile} in the abstract syntax of the DGL language}{%
+languages/fsml/sample-turnstile-dgl.term}{%
+prolog}
+
+The visualization in \autoref{F:turnstile} renders an FSM as a
+graph. There are nodes and edges; both of them can be labeled; edges
+are directed; nodes are of a certain shape (an ellipse here) and style
+(regular or
+filled). \autoref{languages/fsml/sample-turnstile-dgl.term} represents
+represents the graph in the abstract (Prolog-based) syntax of a simple
+graph description based on the DGL notation provided by the \slepro{}
+project. DGL stands for `dot-based graph language'; see \S\ref{A:dgl}
+for this notation. Nodes are quadruplets of the form (\m{Id},
+\m{Label}, \m{Shape}, \m{Style}); edges are triplets of the form
+(\m{FromId}, \m{ToId}, \m{OptionalLabel}).
+
+\codefigure{%
+The FSM of \autoref{F:turnstile} in the dot language}{%
+languages/fsml/sample-turnstile.dot}{%
+dot}
+
+The DGL processor translates the graph into a specification according
+to the \emph{dot} language of
+\emph{GraphViz}\footnote{\url{http://www.graphviz.org/}}; see
+\autoref{languages/fsml/sample-turnstile.dot}. The GraphViz tool
+directly renders the dot specification as shown \autoref{F:turnstile}.
 
@@
 
-TBD
+Let us discuss some characteristics, limitations, or opportunities
+for alternative or additional components. In this manner, we prepare
+for a good number of experiments that could be carried out---either to
+learn more about DSLs or to demonstrate other technologies and
+techniques:
+
+\begin{description}
+
+\item[Leverage a programming ecosystem] The FSML model of
+  this document directly relies on Prolog or language modeling
+  notations that are in turn implemented in Prolog, as part of the
+  \slepro{} project. Alternatively, one could model (implement) FSML
+  also in some existing programming ecosystem, e.g., in Java, Ruby, or
+  Python, also subject to the use of appropriate libraries, parser
+  generators, and other tools. This may enable interesting comparison
+  across programming languages and ecosystems.
+
+\item[Leverage a language workbench] Language modeling is readily
+  addressed by so-called language workbenches such as Rascal or
+  Spoofax. Thus, it would only be natural to exercise such workbenches
+  for the present DSL scenario. In fact, the overall domain of FSMs is
+  generally a popular one, when it comes DSL modeling
+  (implementation). Thus, some similar model may readily exist for
+  some of these workbenches.
+
+\item[Leverage model-driven engineering] Another existing approach to
+  language modeling is based on model-driven engineering (MDE). For
+  instance, one could use EMF and GMF for language modeling, thereby
+  covering abstract and visual syntax for graphical editors. The
+  actual functionality (such as constraints, semantics, and code
+  generator) could leverage a programming language which readily
+  collaborates with the chosen MDE approach, e.g., Java in the case of
+  EMF/GMF. Some of this functionality could also be expressed with
+  model transformations in designated languages, e.g., ATL or
+  QVT. Further, specific model-to-text and text-to-model technologies
+  may be leveraged in the context of MDE.
+
+\item[Concrete object syntax] The FSML model of the present document
+  leverages abstract syntax for all the notations that need to be
+  manipulated: FSML, Java, and DGL. Some metaprogramming systems and
+  language workbenches also support concrete object syntax, e.g.,
+  Stratego and TXL. In this manner, less syntaxes need to be defined,
+  metaprograms remain closer to the notation that one may be used for
+  the manipulated artifacts. In the future, support for concrete
+  object syntax may be added to \slepro.
+
+\item[Template-based code generation] The FSML model underlying the
+  present document leverages pretty-printing combinators for mapping
+  abstract to concrete syntax. In fact, these mapping components are
+  not shown in the document because they are concerned with languages
+  other than FSML, namely Java and DGL. Nevertheless, it may be
+  worthwhile to investigate other options for mapping abstract to
+  concrete syntax---in particular, template-based code generation or
+  model-to-text transformations, as mentioned earlier in the context
+  of MDE.
+
+\item[Alternative code generation schemes] The code generator favors
+  one particular scheme such that FSMs are essentially represented as
+  data and actions are to be plugged into FSM execution through
+  appropriate handlers. Several other schemes for the OO
+  representation of FSMs are known. For instance, transitions may also
+  be represented as control-flow code, i.e., by using if or switch
+  statements dispatching on states and inputs. Alternatively, states
+  may also be thought of as objects with transitions modeled as
+  polymorphic methods. Without caring much here about the pros and
+  cons of these approaches, we could simply want to study them from
+  the technical point of view of code generation.  In fact, we picked
+  the data-centric approach, as it was obvious that code generation is
+  straightforward in this case.
+
+\item[An editor for FSML] Language support may also include editing
+  support as in the sense of a structure editor for concrete textual
+  syntax or a graphical editor for concrete visual syntax (perhaps
+  based on GMF, as mentioned earlier).
+
+\item[A proper FSML-based application] The value of the FSML language
+  may be more obvious, if we managed to demonstrate it in an actual
+  application such as a concrete turnstile system. One could be
+  looking at embedded systems scenarios, for example.
+
+\end{description}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\appendix
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\pagebreak
+
+\section{Grammar of EGL grammars}
+\label{A:egl}
+
+\codefigure{%
+EGL grammar of EGL grammars}{%
+languages/egl/cs.egl}{%
+egl}
+
+EGL was used for the concrete syntax definition of FSML in
+\S\ref{S:textual}. EGL is an EBNF-like notation. See
+\autoref{languages/egl/cs.egl} for a self-description of the grammar notation.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\pagebreak
+
+\section{Grammar of ESL signature}
+\label{A:esl}
+
+\codefigure{%
+EGL grammar of ESL signatures}{%
+languages/esl/cs.egl}{%
+esl}
+
+ESL was used for the abstract syntax definition of FSML in
+\S\ref{S:abstract}. ESL is a type-declaration notation inspired by
+algebraic signatures, term algebras, and algebraic data types.  See
+\autoref{languages/esl/cs.egl} for the concrete syntax of the
+notation. ESL is also used in some of the appendix sections that follow.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\pagebreak
+
+\section{Abstract syntax of a Java subset}
+\label{A:java}
+
+\codefigure{%
+ESL signature for the abstract syntax of a Java subset}{%
+languages/java/as.esl}{%
+esl}
+
+See \autoref{languages/java/as.esl}. The subset is assumed by the code
+generator of \S\ref{S:generator}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\pagebreak
+
+\section{Abstract syntax of DGL}
+\label{A:dgl}
+
+\codefigure{%
+ESL signature for the abstract syntax of DGL}{%
+languages/dgl/as.esl}{%
+esl}
+
+We used DGL in \S\ref{S:visual} to export FSMs as graphs in a manner
+that they can be visualized. DGL is a simple graph description
+language. DGL stands `dot-based graph language' to hint at the fact
+that the abstract syntax of DGL can be (is) rendered in the dot
+language of Graphviz. The abstract syntax of DGL is shown in
+\autoref{languages/dgl/as.esl}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \end{document}

### 2013-11-09 languages/fsml/java/Stepper.java

@@ -5,22 +5,21 @@ public abstract class Stepper<S, I, A> {
     protected S state;
     protected Handler<A> handler;
     private HashMap<S,HashMap<I,Pair<A,S>>> table =
-	new HashMap<S,HashMap<I,Pair<A,S>>>();
+        new HashMap<S,HashMap<I,Pair<A,S>>>();
     public final void add(S from, I i, A a, S to) {
-	if (!table.containsKey(from))
-	    table.put(from, new HashMap<I,Pair<A,S>>());
-	HashMap<I,Pair<A,S>> subtable = table.get(from);
-	Pair<A,S> pair = new Pair<A,S>(a, to);
-	subtable.put(i, pair);
+        if (!table.containsKey(from))
+            table.put(from, new HashMap<I,Pair<A,S>>());
+        HashMap<I,Pair<A,S>> subtable = table.get(from);
+        Pair<A,S> pair = new Pair<A,S>(a, to);
+        subtable.put(i, pair);
     }
     public final void step(I i) {
-	HashMap<I,Pair<A,S>> subtable = table.get(state);
-	Pair<A,S> pair = subtable.get(i);
-	S from = state;
+        HashMap<I,Pair<A,S>> subtable = table.get(state);
+        Pair<A,S> pair = subtable.get(i);
+        S from = state;
         S to = pair.y;
-	System.out.println("from: "+from+", input: "+i+", to: "+to);
-	if (pair.x!=null)
-	    handler.handle(pair.x);
-	state = to;
+        System.out.println("from: "+from+", input: "+i+", to: "+to);
+        if (pair.x!=null) handler.handle(pair.x);
+        state = to;
     }
 }

### 2013-11-09 languages/fsml/java/TurnstileHandler.java

@@ -1,16 +1,16 @@
 // Turnstile-specific, handwritten code 
 public class TurnstileHandler implements Handler<Action> {
     public void handle(Action a) {
-	switch (a) {
-	case collect : 
-	    System.out.println("collect");
-	    break;
-	case alarm : 
-	    System.out.println("alarm");
-	    break;
-	case eject : 
-	    System.out.println("eject");
-	    break;
-	}
+        switch (a) {
+        case collect : 
+            System.out.println("collect");
+            break;
+        case alarm : 
+            System.out.println("alarm");
+            break;
+        case eject : 
+            System.out.println("eject");
+            break;
+        }
     }
 }

### 2013-11-12 docs/fsml/paper.tex
Made the accept predicates a bit more regular

@@
 ad-hoc grammar notation that is provided by the \slepro{} project; see
-\S\ref{A:egl} for a self-description. In particular, EGL grammars can
+\ref{A:egl} for a self-description. In particular, EGL grammars can
 be executed for the purpose of parsing, but we omit all such details
@@
 programming. ESL is an ad-hoc signature notation that is provided by
-the \slepro{} project; see \S\ref{A:esl} for a syntax definition of
+the \slepro{} project; see \ref{A:esl} for a syntax definition of
 the signature notation. In particular, ESL grammars can be executed
@@
 ids and assembles an enum type, subject to a suitable abstract syntax
-of Java; see \S\ref{A:java} for a summary of the (abstract) Java
+of Java; see \ref{A:java} for a summary of the (abstract) Java
 syntax used by the code generator. Finally, the Java declaration is
@@
 graph description based on the DGL notation provided by the \slepro{}
-project. DGL stands for `dot-based graph language'; see \S\ref{A:dgl}
+project. DGL stands for `dot-based graph language'; see \ref{A:dgl}
 for this notation. Nodes are quadruplets of the form (\m{Id},

### 2013-11-26 docs/fsml/paper.tex
Started to use "megamodel" patterns.

@@
 
-0.00001 as of 8 November 2013.
+0.00002 as of 25 November 2013.
 
@@
 \vspace{-77\in}
-\includegraphics[width=\textwidth]{../../languages/fsml/sample-turnstile.pdf}
+\includegraphics[width=\textwidth]{../../languages/fsml/sample.pdf}
 \vspace{-150\in}
@@
 The FSM of \autoref{F:turnstile} in concrete textual syntax}{%
-languages/fsml/sample-turnstile.fsml}{%
+languages/fsml/sample.fsml}{%
 fsml}
@@
 mean that the transition's target is the current state. See
-\autoref{languages/fsml/sample-turnstile.fsml} for an illustration.
+\autoref{languages/fsml/sample.fsml} for an illustration.
 
@@
 assumed that the concrete textual syntax, as illustrated in
-\autoref{languages/fsml/sample-turnstile.fsml}, can be parsed indeed
+\autoref{languages/fsml/sample.fsml}, can be parsed indeed
 with the present grammar.
@@
 The FSM of \autoref{F:turnstile} in abstract syntax}{%
-languages/fsml/sample-turnstile-fsml.term}{%
+languages/fsml/sample.term}{%
 prolog}
 
-See \autoref{languages/fsml/sample-turnstile-fsml.term} for an
+See \autoref{languages/fsml/sample.term} for an
 illustration. The abstract syntax describes a FSM as a list of state
@@
 \autoref{languages/fsml/as.esl} for the signature. The term in
-\autoref{languages/fsml/sample-turnstile-fsml.term} conforms indeed to
+\autoref{languages/fsml/sample.term} conforms indeed to
 the present signature.
@@
 The FSM of \autoref{F:turnstile} in the abstract syntax of the DGL language}{%
-languages/fsml/sample-turnstile-dgl.term}{%
+languages/fsml/sample.dgl}{%
 prolog}
@@
 (regular or
-filled). \autoref{languages/fsml/sample-turnstile-dgl.term} represents
+filled). \autoref{languages/fsml/sample.dgl} represents
 represents the graph in the abstract (Prolog-based) syntax of a simple
@@
 The FSM of \autoref{F:turnstile} in the dot language}{%
-languages/fsml/sample-turnstile.dot}{%
+languages/fsml/sample.dot}{%
 dot}
@@
 \emph{GraphViz}\footnote{\url{http://www.graphviz.org/}}; see
-\autoref{languages/fsml/sample-turnstile.dot}. The GraphViz tool
+\autoref{languages/fsml/sample.dot}. The GraphViz tool
 directly renders the dot specification as shown \autoref{F:turnstile}.

### 2013-11-26 docs/ueber/paper.tex
Some incremental improvements

--- 2013-11-26 docs/fsml/paper.tex
@@
 \newcommand{\ooo}[1]{\textsf{101#1}}
+\newcommand{\ueber}{\textsc{ueber}}
 
@@
 
-\title{Another DSL primer}
+\title{The megamodeling language \ueber}
 
-\author{Ralf L\"ammel}
+\author{Ralf L\"ammel and Andrei Varanovich}
 
@@
 \begin{abstract}
-  We model a domain-specific language FSML for Finite State Machines
-  (FSMs). The language model encompasses concrete a textual syntax, a
-  term-based abstract syntax, a graph-based visual syntax, extra
-  constraints for well-formedness, a simulation semantics, and a
-  code-generation semantics for representing and executing FSMs in
-  Java. The key motivation for the present model of FSML to be
-  explainable exhaustively in terms of simple formalisms and idioms,
-  all layered on top of Prolog. The FSML development is part of the
-  \slepro{} project; see \url{https://github.com/slebok/slepro}.
+  We describe the megamodeling language \ueber, as it is used for
+  modeling the linguistic architecture in the \slepro{} project; see
+  \url{https://github.com/slebok/slepro}.
 
@@
 
+\begin{comment}
 \noindent
@@
   gratefully acknowledged.}
+\end{comment}
 
@@
 
-0.00002 as of 25 November 2013.
+0.00001 as of 26 November 2013.
 
@@
 
-\url{https://github.com/slebok/slepro/tree/master/docs/fsml}.
+\url{https://github.com/slebok/slepro/tree/master/docs/ueber}.
 \end{abstract}
@@
 
-\section{Introducing FSML}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[t!]
-\vspace{-77\in}
-\includegraphics[width=\textwidth]{../../languages/fsml/sample.pdf}
-\vspace{-150\in}
-\caption{A finite state machine for turnstiles.}
-\label{F:turnstile}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Consider the visual representation of a finite state machine (FSM) for
-a turnstile (as used in subway systems). The idea is, of course, that
-the customer is supposed to insert a ticket before passing the
-turnstile. Thus, the FSM assumes the initial state `locked'. (The
-initial state is highlighted by using a filled ellipse.) The input
-`ticket' (meaning the insertion of a ticket) causes a transition to
-the state `unlocked' and the action `collect'. In this new state, the
-input `pass' (meaning the attempt to pass the turnstile) causes a
-transition back to the state `locked'. If the input `pass' was made in
-the state `locked', then this causes a transition to the state
-`exception' and an action `alarm'. The state `exception' rejects the
-input `ticket'. The input `pass' does, of course, keeps one in the
-state `exception'. The `alarm' can be muted with the input `mute'. We
-return to the normal `locked' state upon the input `release'.
-
-We will model a language FSML (FSM language) for modeling
-FSMs. Specifically, we will model several aspects of FSML:
-
-\begin{itemize}
-\item The visual syntax as used in the figure; see \S\ref{S:visual}.
-\item A textual syntax as an alternative concrete syntax; see \S\ref{S:textual}.
-\item An abstract syntax useful for representation in programs; see \S\ref{S:abstract}.
-\item Extra well-formedness constraints on the abstract syntax; see \S\ref{S:ok}.
-\item A reference semantics for the simulation of FSMs; see \S\ref{S:semantics}.
-\item A code generator translating FSMs into OO programs; see \S\ref{S:generator}.
-\end{itemize}
-
-This primer is concluded by some `food for thought' in
-\S\ref{S:concl}. That is, we critically discuss the language model at
-hand as well as the underlying approach, thereby also motivating
-reproductions of the model as well as variations on the model,
-possibly leveraging diverse technologies and techniques.
-
-The FSML development is part of the \slepro{} project. All software
-artifacts of this document and all support infrastructure is available
-through the repository of the project:
-
-\begin{center}
-\url{https://github.com/slebok/slepro}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{The concrete textual syntax of FSML}
-\label{S:textual}
-
-\codefigure{%
-The FSM of \autoref{F:turnstile} in concrete textual syntax}{%
-languages/fsml/sample.fsml}{%
-fsml}
-
-A textual syntax may break down an FSM into `state declarations',
-i.e., states with the associated transitions. Each transition
-identifies the relevant input, the optional action, and the target
-state. In fact, if the target state is omitted, then this is taken to
-mean that the transition's target is the current state. See
-\autoref{languages/fsml/sample.fsml} for an illustration.
-
-\codefigure{%
-Context-free grammar for the concrete syntax of FSML}{%
-languages/fsml/cs.egl}{%
-egl}
-
-Let us define the syntax in terms of a context-free grammar. We use an
-EGL grammar, where EGL stands for `extended grammar language' and can
-be regarded as a variation on the Extended Backus Naur Form. EGL is an
-ad-hoc grammar notation that is provided by the \slepro{} project; see
-\ref{A:egl} for a self-description. In particular, EGL grammars can
-be executed for the purpose of parsing, but we omit all such details
-here. See \autoref{languages/fsml/cs.egl} for the grammar. It can be
-assumed that the concrete textual syntax, as illustrated in
-\autoref{languages/fsml/sample.fsml}, can be parsed indeed
-with the present grammar.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{The abstract syntax of FSML}
-\label{S:abstract}
-
-Eventually, we want to manipulate FSMs in (Prolog) programs. To this
-end, we need a suitable abstract syntax. As an aside, there exist
-metaprogramming systems that can leverage the concrete syntax
-directly, without any need for a separate abstract syntax, but we will
-adopt the simpler approach here to indeed use an abstract syntax. In
-fact, we assume that abstract syntax trees are (Prolog) terms as in
-the sense of term algebras or algebraic specifications.
-
-\codefigure{%
-The FSM of \autoref{F:turnstile} in abstract syntax}{%
-languages/fsml/sample.term}{%
-prolog}
-
-See \autoref{languages/fsml/sample.term} for an
-illustration. The abstract syntax describes a FSM as a list of state
-declarations, which are in turn tuples (triplets) of the form (\m{V},
-\m{Id}, \m{Ts}) where \m{V} is a Boolean saying whether the state is
-initial, \m{Id} is the assigned id (name) of the state, and \m{Ts} is
-a list of transitions. Each transition is a triplet of the form
-(\m{In}, \m{A}, \m{To}) where \m{In} is the input for the transition,
-\m{A} is the optional action and \m{To} is the target state. The
-optionality of the action is represented such that the missing action
-is represented by `[]' and the present action, e.g., `eject', is
-represented by `[eject]'.
-
-\codefigure{%
-Signature for the abstract syntax of FSML}{%
-languages/fsml/as.esl}{%
-esl}
-
-Let us define the abstract syntax in terms of a signature that defines
-all valid FSM terms. We use an ESL signature, where ESL stands for
-`extended signature language' and can be regarded as (quite) a
-variation on notations familiar from algebraic specification or (less)
-a variation on type systems used in logic or functional
-programming. ESL is an ad-hoc signature notation that is provided by
-the \slepro{} project; see \ref{A:esl} for a syntax definition of
-the signature notation. In particular, ESL grammars can be executed
-for conformance checking to see whether a given term conforms to a
-given signature, but we omit all such details here. See
-\autoref{languages/fsml/as.esl} for the signature. The term in
-\autoref{languages/fsml/sample.term} conforms indeed to
-the present signature.
-
-\codefigure{%
-Concrete to abstract syntax mapping for FSML}{%
-languages/fsml/cs-to-as.pro}{%
-prolog}
-
-As we intend to parse FSMs via the concrete textual syntax and to
-manipulate FSMs though via the abstract term-based syntax, we need a
-mapping from the former to the latter. Of course, both syntaxes are
-very close to each other in terms of the nonterminals or types defined
-and in terms of the structural breakdown. However, some fine details
-need to be declared explicitly via a mapping so that parse trees
-according to the earlier grammar are precisely mapped to terms that
-conform to the signature for the abstract syntax. The mapping is
-defined by a Prolog predicate \emph{fsmlMapping}; see the Prolog
-clauses in \autoref{languages/fsml/cs-to-as.pro}.
-
-The syntax definition approach of the \slepro{} project assumes such
-mappings are indeed predicates with three arguments \m{N}, \m{T1},
-\m{T2} as follows. \m{N} is the nonterminal (according to the grammar
-for the concrete syntax) whose parse trees are to be mapped, i.e.,
-rewritten. \m{T1} is the (imploded) parse tree that just
-systematically follows the structure of the grammar. \m{T2} is the
-term that should be used in place of \m{T1}. The starting point is an
-identity mapping; the mapping predicate only overwrites those cases
-where the identity mapping is not appropriate. In
-\autoref{languages/fsml/cs-to-as.pro}, the first clause replaces the
-list-of-chars representation of FSM names by proper atoms (`strings').
-The second clause replaces `initial' by `true' (and `noninitial' by
-`false'); it also fills in the target state, when it is missing,
-thereby taking care of some syntactic sugar, i.e., the permitted
-omission of the target state when it equals the source state of a
-transition.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Constraints on the abstract syntax}
-\label{S:ok}
-
-The concrete and abstract syntaxes so far do not cover some of the
-constraints that we would naturally apply to FSMs. For instance, we
-did not rule out so far that FSMs have multiple initial states, which
-is probably not useful. Context-free grammars and signatures are
-notoriously limited in expressiveness to model such constraints,
-whereas they are convenient for the basic definition of structure. We
-need to add extra constraints on top of (say) the abstract syntax
-definition. In some communities, it is common to actually consider
-these constraints to form part of the abstract syntax whereas we
-follow another common view that such constraints are considered
-separately as in well-formedness or well-typedness judgements used in
-type systems for programming languages.
-
-\codefigure{%
-Overview of constraints imposed on FSML's abstract syntax}{%
-languages/fsml/ok.pro}{%
-prolog}
-
-See \autoref{languages/fsml/ok.pro} for the Prolog specification
-listing the named constraints that we consider here. Each constraint
-gives rise to a separate predicate considered below. Here is short
-summary. Constraint \emph{fsmSingleInitial} is meant to ensure that
-there is only a single state declaration for an initial
-state. Constraint \emph{fsmDistinctIds} is meant to ensure that the
-state ids of the state declarations are distinct. Constraint
-\emph{fsmResolvable} is meant to ensure that all referenced target
-states are declared. Constraint \emph{fsmDeterministic} is meant
-to ensure that each given input uniquely defines the target state to
-be transitioned to, if any. Constraint \emph{fsmReachable} is meant to
-ensure that all declared states are reachable from the
-initial state. In \autoref{languages/fsml/ok.pro}, all constraint
-predicates are surround by an application of the meta-predicate
-\emph{require/1}, which is simply there for better error
-reporting. That is, if the argument of \emph{require/1} fails, then
-this is reported immediately, as opposed to simply propagating failure
-silently.
-
-\codefigure{%
-There is only a single state declaration for an initial state}{%
-languages/fsml/ok/initial.pro}{%
-prolog}
-
-\codefigure{%
-The state ids of the state declarations are distinct}{%
-languages/fsml/ok/distinct.pro}{%
-prolog}
-
-\codefigure{%
-All referenced target states are declared}{%
-languages/fsml/ok/resolvable.pro}{%
-prolog}
-
-\codefigure{%
-Each given input uniquely defines the target state to be transitioned to, if any}{%
-languages/fsml/ok/deterministic.pro}{%
-prolog}
-
-\codefigure{%
-All declared states are reachable from the initial state}{%
-languages/fsml/ok/reachable.pro}{%
-prolog}
-
-\codefigure{%
-Fixed point computation needed in \autoref{languages/fsml/ok/reachable.pro}}{%
-languages/fsml/ok/closure.pro}{%
-prolog}
-
-All constraints are specified in
-\autoref{languages/fsml/ok/initial.pro}--\autoref{languages/fsml/ok/closure.pro}. The
-constraints are relatively simple to express using Prolog's
-meta-predicate \emph{findall/3} which serves here for queries into the
-terms of the abstract syntax. For instance, in
-\autoref{languages/fsml/ok/initial.pro}, we find all state ids from
-state declarations with `true' for the initial component. Once this
-set is retrieved that the resulting list is of length 1; thus, there
-is exactly one initial.
-
-The constraint for reachability
-(\autoref{languages/fsml/ok/reachable.pro}--\autoref{languages/fsml/ok/closure.pro})
-is somewhat interesting, as we have to model a fixed point computation
-to find all states by repeated consideration of transition as to
-whether the set of known to be reachable states implies more reachable
-states. The fixed point criterion is here that we do not find
-additional states; see \autoref{languages/fsml/ok/closure.pro}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{The reference semantics for FSML}
-\label{S:semantics}
-
-So far we relies on an intuitive understanding of the FSML semantics,
-which is reasonable in so far that the notion of finite state machines
-is pretty fundamental and familiar. Nevertheless, we are bound to
-provide a precise and executable reference semantics of the language
-which can then be hold accountable by other language-based software
-components, e.g., a code generator (see next section) or a
-refactoring.
-
-We speak of a reference semantics here to emphasize that this
-semantics may not be practically useful immediately. That is, we
-will define the semantics of a FSM by a simulation judgement which
-assumes that all input is provided upfront and all output (the action
-sequence) is provided upon completion of input processing. This is a
-batch model, which is useful in a formal semantics, but it is not
-useful for using FSMs in interactive systems.
-
-\codefigure{%
-Semantics of FSML}{%
-languages/fsml/simulation.pro}{%
-prolog}
-
-See \autoref{languages/fsml/simulation.pro} for the semantics; the
-judgements are expressed in Prolog. The clause at the top selects the
-initial state of the FSM and enters the simulation with that state and
-the complete input. The remaining two clauses describe the iteration
-over the input such that a suitable transition is looked up in each
-step and action (if any) and the new state are paired up as an element in
-the output so that simulation proceeds with the new state and the
-remaining input.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{The code generator for FSML}
-\label{S:generator}
-
-A practically useful interpretation of FSML would be achieved, if we
-enabled interactive execution of the FSMs in the context of some
-programming environment. We pick Java as our target here. We describe
-a code generator which translates FSMs into Java code. The generated
-code can be augmented with hand-written components specifically for
-the intended semantics of actions. The Java representation of FSMs
-meets the requirement that execution is step-by-step. That is, a
-transition is transparently realized including the potential
-performance of an action whenever an input item arrives.
-
-There are many different ways to generate imperative/OO code for
-FSMs. We pick one particular option here--without loss of
-generality. The chosen option can be said to be data-centric in that
-FSMs are represented in an appropriate container-based data structure,
-whereas the execution of FSMs relies on the interpretation of said
-data structure. Further, the programmer-defined meaning of actions is
-provided by a handler which interpreters action names.
-
-\codefigure{%
-Java representation of the states for the turnstile example}{%
-languages/fsml/java/State.java}{%
-java}
-
-\codefigure{%
-Java representation of the inputs for the turnstile example}{%
-languages/fsml/java/Input.java}{%
-java}
-
-\codefigure{%
-Java representation of the actions for the turnstile example}{%
-languages/fsml/java/Action.java}{%
-java}
-
-\codefigure{%
-Code generation for state type}{%
-languages/fsml/to-java/state.pro}{%
-prolog}
-
-\autoref{languages/fsml/java/State.java}--\autoref{languages/fsml/java/Action.java}
-shows (generated) Java enum types for the states, inputs, and
-actions. Obviously, these sets are trivially obtainable from the
-abstract syntactical representation of an FSM. In
-\autoref{languages/fsml/to-java/state.pro}, this is illustrated for
-the enum type for states. The predicate collects all (declared) state
-ids and assembles an enum type, subject to a suitable abstract syntax
-of Java; see \ref{A:java} for a summary of the (abstract) Java
-syntax used by the code generator. Finally, the Java declaration is
-pretty printed; see the use of \emph{ppJavaDecl/2} in the figure,
-subject to a pretty printer for the Java language. The generator
-predicates for the enum types for inputs and actions are quite
-similar.
-
-\codefigure{%
-Generic interface for action handers}{%
-languages/fsml/java/Handler.java}{%
-java}
-
-\codefigure{%
-A trivial interpretation of the turnstile actions}{%
-languages/fsml/java/TurnstileHandler.java}{%
-java}
-
-\autoref{languages/fsml/java/Handler.java} shows a generic, i.e.,
-FSM-independent interface for defining handlers for FSM
-actions. \autoref{languages/fsml/java/TurnstileHandler.java} shows a
-trivial implementation of said interface for the turnstile
-example. The actions simply print out the names of the actions, but it
-is clear that arbitrary functionality can be plugged into FSM
-execution in this manner.
-
-\codefigure{%
-Generic class for FSM execution}{%
-languages/fsml/java/Stepper.java}{%
-java}
-
-\autoref{languages/fsml/java/Stepper.java} shows essentially the
-runtime for FSMs. This is generic code which is appropriately
-parameterized in the types for states, inputs, and actions. The class
-relies on a table for holding the transitions in the form of a map
-from states to maps from inputs to pairs of optional actions and
-states. There is an appropriate \emph{add} method which makes it easy
-to insert into this non-basic container. The class is also prepared to
-keep track of and make use of a handler for the involved actions. The
-class is called \emph{Stepper} because its key method is the one for
-performing a step (a transition) based on an actual input item.
-
-\codefigure{%
-Configuration of a stepper for the turnstile example}{%
-languages/fsml/java/TurnstileStepper.java}{%
-java}
-
-\autoref{languages/fsml/java/TurnstileStepper.java} shows the
-generated code for the turnstile-specific \emph{Stepper} subclass. The
-class instantiates the generic parameters of the \emph{Stepper} class
-to the turnstile-specific enum types for states, inputs, and actions,
-which were shown earlier. Also, the class encodes all transitions of
-the turnstile FSM via appropriate calls to the \emph{add} method, as
-part of the constructor. The initial state is also set up and the
-turnstile-specific handler is communicated as well.
-
-\codefigure{%
-Code generation for stepper subclass}{%
-languages/fsml/to-java/stepper.pro}{%
-prolog}
-
-\autoref{languages/fsml/to-java/stepper.pro} shows the code generation
-predicate for stepper subclasses. The predicate refers to more
-elements of the Java language: several expression forms (e.g.,
-`null'), assignment statements, method calls, and class
-declarations. In a final step, the constructed abstract syntactical
-representation is passed to the pretty printer for Java.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{The visual syntax of FSML}
-\label{S:visual}
-
-It remains to define the visual syntax of FSML, as it was illustrated
-in \autoref{F:turnstile}. A visual syntax definition may serve these
-major purposes: actual representation (visualization) and
-editing. Indeed, we want this syntax to be complemented by rendering
-support. We do not consider editing here.
-
-\codefigure{%
-The FSM of \autoref{F:turnstile} in the abstract syntax of the DGL language}{%
-languages/fsml/sample.dgl}{%
-prolog}
-
-The visualization in \autoref{F:turnstile} renders an FSM as a
-graph. There are nodes and edges; both of them can be labeled; edges
-are directed; nodes are of a certain shape (an ellipse here) and style
-(regular or
-filled). \autoref{languages/fsml/sample.dgl} represents
-represents the graph in the abstract (Prolog-based) syntax of a simple
-graph description based on the DGL notation provided by the \slepro{}
-project. DGL stands for `dot-based graph language'; see \ref{A:dgl}
-for this notation. Nodes are quadruplets of the form (\m{Id},
-\m{Label}, \m{Shape}, \m{Style}); edges are triplets of the form
-(\m{FromId}, \m{ToId}, \m{OptionalLabel}).
-
-\codefigure{%
-The FSM of \autoref{F:turnstile} in the dot language}{%
-languages/fsml/sample.dot}{%
-dot}
-
-The DGL processor translates the graph into a specification according
-to the \emph{dot} language of
-\emph{GraphViz}\footnote{\url{http://www.graphviz.org/}}; see
-\autoref{languages/fsml/sample.dot}. The GraphViz tool
-directly renders the dot specification as shown \autoref{F:turnstile}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Food for thought}
-\label{S:concl}
-
-Let us discuss some characteristics, limitations, or opportunities
-for alternative or additional components. In this manner, we prepare
-for a good number of experiments that could be carried out---either to
-learn more about DSLs or to demonstrate other technologies and
-techniques:
-
-\begin{description}
-
-\item[Leverage a programming ecosystem] The FSML model of
-  this document directly relies on Prolog or language modeling
-  notations that are in turn implemented in Prolog, as part of the
-  \slepro{} project. Alternatively, one could model (implement) FSML
-  also in some existing programming ecosystem, e.g., in Java, Ruby, or
-  Python, also subject to the use of appropriate libraries, parser
-  generators, and other tools. This may enable interesting comparison
-  across programming languages and ecosystems.
-
-\item[Leverage a language workbench] Language modeling is readily
-  addressed by so-called language workbenches such as Rascal or
-  Spoofax. Thus, it would only be natural to exercise such workbenches
-  for the present DSL scenario. In fact, the overall domain of FSMs is
-  generally a popular one, when it comes DSL modeling
-  (implementation). Thus, some similar model may readily exist for
-  some of these workbenches.
-
-\item[Leverage model-driven engineering] Another existing approach to
-  language modeling is based on model-driven engineering (MDE). For
-  instance, one could use EMF and GMF for language modeling, thereby
-  covering abstract and visual syntax for graphical editors. The
-  actual functionality (such as constraints, semantics, and code
-  generator) could leverage a programming language which readily
-  collaborates with the chosen MDE approach, e.g., Java in the case of
-  EMF/GMF. Some of this functionality could also be expressed with
-  model transformations in designated languages, e.g., ATL or
-  QVT. Further, specific model-to-text and text-to-model technologies
-  may be leveraged in the context of MDE.
-
-\item[Concrete object syntax] The FSML model of the present document
-  leverages abstract syntax for all the notations that need to be
-  manipulated: FSML, Java, and DGL. Some metaprogramming systems and
-  language workbenches also support concrete object syntax, e.g.,
-  Stratego and TXL. In this manner, less syntaxes need to be defined,
-  metaprograms remain closer to the notation that one may be used for
-  the manipulated artifacts. In the future, support for concrete
-  object syntax may be added to \slepro.
-
-\item[Template-based code generation] The FSML model underlying the
-  present document leverages pretty-printing combinators for mapping
-  abstract to concrete syntax. In fact, these mapping components are
-  not shown in the document because they are concerned with languages
-  other than FSML, namely Java and DGL. Nevertheless, it may be
-  worthwhile to investigate other options for mapping abstract to
-  concrete syntax---in particular, template-based code generation or
-  model-to-text transformations, as mentioned earlier in the context
-  of MDE.
-
-\item[Alternative code generation schemes] The code generator favors
-  one particular scheme such that FSMs are essentially represented as
-  data and actions are to be plugged into FSM execution through
-  appropriate handlers. Several other schemes for the OO
-  representation of FSMs are known. For instance, transitions may also
-  be represented as control-flow code, i.e., by using if or switch
-  statements dispatching on states and inputs. Alternatively, states
-  may also be thought of as objects with transitions modeled as
-  polymorphic methods. Without caring much here about the pros and
-  cons of these approaches, we could simply want to study them from
-  the technical point of view of code generation.  In fact, we picked
-  the data-centric approach, as it was obvious that code generation is
-  straightforward in this case.
-
-\item[An editor for FSML] Language support may also include editing
-  support as in the sense of a structure editor for concrete textual
-  syntax or a graphical editor for concrete visual syntax (perhaps
-  based on GMF, as mentioned earlier).
-
-\item[A proper FSML-based application] The value of the FSML language
-  may be more obvious, if we managed to demonstrate it in an actual
-  application such as a concrete turnstile system. One could be
-  looking at embedded systems scenarios, for example.
-
-\end{description}
+\section{???}
 
@@
 
-\appendix
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\pagebreak
-
-\section{Grammar of EGL grammars}
-\label{A:egl}
-
-\codefigure{%
-EGL grammar of EGL grammars}{%
-languages/egl/cs.egl}{%
-egl}
-
-EGL was used for the concrete syntax definition of FSML in
-\S\ref{S:textual}. EGL is an EBNF-like notation. See
-\autoref{languages/egl/cs.egl} for a self-description of the grammar notation.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\pagebreak
-
-\section{Grammar of ESL signature}
-\label{A:esl}
-
-\codefigure{%
-EGL grammar of ESL signatures}{%
-languages/esl/cs.egl}{%
-esl}
-
-ESL was used for the abstract syntax definition of FSML in
-\S\ref{S:abstract}. ESL is a type-declaration notation inspired by
-algebraic signatures, term algebras, and algebraic data types.  See
-\autoref{languages/esl/cs.egl} for the concrete syntax of the
-notation. ESL is also used in some of the appendix sections that follow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\pagebreak
-
-\section{Abstract syntax of a Java subset}
-\label{A:java}
-
-\codefigure{%
-ESL signature for the abstract syntax of a Java subset}{%
-languages/java/as.esl}{%
-esl}
-
-See \autoref{languages/java/as.esl}. The subset is assumed by the code
-generator of \S\ref{S:generator}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\pagebreak
-
-\section{Abstract syntax of DGL}
-\label{A:dgl}
-
-\codefigure{%
-ESL signature for the abstract syntax of DGL}{%
-languages/dgl/as.esl}{%
-esl}
-
-We used DGL in \S\ref{S:visual} to export FSMs as graphs in a manner
-that they can be visualized. DGL is a simple graph description
-language. DGL stands `dot-based graph language' to hint at the fact
-that the abstract syntax of DGL can be (is) rendered in the dot
-language of Graphviz. The abstract syntax of DGL is shown in
-\autoref{languages/dgl/as.esl}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2013-11-28 docs/ueber/paper.tex
Intermediate commit; ueber is broken right now.

@@
 \usepackage{amssymb}
-\usepackage{listings,lsthaskell,lstprolog,lstegl,lstesl,lstfsml,lstjava,lstdot}
+\usepackage{listings,lsthaskell,lstprolog,lstegl,lstesl,lstfsml,lstjava,lstdot,lsttext,lstueber}
 
@@
 \newcommand{\ooo}[1]{\textsf{101#1}}
-\newcommand{\ueber}{\textsc{ueber}}
+\newcommand{\ueber}{\textsf{ueber}}
+\newcommand{\ppl}{\textsf{ppl}}
+\newcommand{\esl}{\textsf{esl}}
 
@@
   We describe the megamodeling language \ueber, as it is used for
-  modeling the linguistic architecture in the \slepro{} project; see
-  \url{https://github.com/slebok/slepro}.
+  modeling the linguistic architecture of the Prolog-based \slepro{}
+  project; see \url{https://github.com/slebok/slepro}.
 
@@
 
-\section{???}
+\section{Introducing \ueber}
+
+The notion of megamodel has been defined and used in different ways in
+software engineering. We favor here the view that a megamodel models
+the architecture of a software system with particular focus on the
+involved languages, technologies, and general software
+concepts~\citep{FavreLV12}
+
+In this document, we present a particular megamodeling language,
+\ueber, which is used for megamodeling the Prolog-based \slepro{}
+project, which is essentially a repository of language modeling
+artifacts: parsers, translators, semantics definitions, etc.
+
+In the following example, we demonstrate \ueber{} in a way as if we
+were testing a language processor. Megamodeling is in no way focused
+or restricted to software testing, but it happens that the particular
+setup of the \slepro{} repository lets megamodeling be useful for
+testing.
+
+\codefigure{%
+A \ppl{} expresion composing text literals vertically}{%
+languages/ppl/test/vbox.ppl}{%
+prolog}
+
+\codefigure{%
+The text rendered for the \ppl{} expression of \autoref{languages/ppl/test/vbox.ppl}}{%
+languages/ppl/test/vbox.txt}{%
+text}
+
+We are concerned with a simple language \ppl---the Pretty Printing
+Language.\footnote{\url{https://github.com/slebok/slepro/tree/master/languages/ppl}}
+This language offers a number of simple composition combinators for
+composing text vertically and horizontally. Consider the \ppl{}
+expression of \autoref{languages/ppl/test/vbox.ppl}. \ppl{} comes with
+a simple evaluation semantics that essentially renders the \ppl{}
+expressions; see \autoref{languages/ppl/test/vbox.txt}.
+
+\codefigure{%
+Evaluation rule for vertical boxes}{%
+languages/ppl/eval/vbox.pro}{%
+prolog}
+
+\codefigure{%
+Final rendering step: turn lists of lines into a single string}{%
+languages/ppl/to-string.pro}{%
+prolog}
+
+\autoref{languages/ppl/eval/vbox.pro} shows one rule of the evaluator
+for \ppl{} terms---the rule for vertical
+composition. \autoref{languages/ppl/to-string.pro} shows the trivial
+conversion from lists of lines to a single string. This is a
+straigthforward Prolog-based implementation of the pretty printing
+combinators. The details do not matter much to the megamodeling
+discussion that follows.
+
+\codefigure{%
+Syntax of \ppl{} terms}{%
+languages/ppl/as.esl}{%
+esl}
+
+The syntax of all \ppl{} terms is defined in
+\autoref{languages/ppl/as.esl}. We use a simple language, \esl, for
+term algebras or algebraic data types, which we do not discuss here in
+detail. The notation should be straightforward; we declare a number of
+different symbols (`constructors') for different kinds of \ppl{}
+combinators. The sample term of \autoref{languages/ppl/test/vbox.ppl}
+conforms to the signature.
+
+\codefigure{%
+A megamodel for testing \ueber}{%
+languages/ppl/test/.ueber}{%
+ueber}
 
@@
 
-%\nocite{FreeDictionary}
 %\setlength{\bibsep}{4pt}
-%\bibliographystyle{elsarticle-harv}
+\bibliographystyle{elsarticle-harv}
 %\bibliographystyle{elsarticle-num}
-%\bibliography{paper}
+\bibliography{paper}
 

### 2013-11-28 languages/fsml/java/Demo.java

@@ -1,10 +1,10 @@
 // Turnstile-specific, handwritten code
-public class TurnstileDemo {
+public class Demo {
     public static void main(String [] args) {
-	TurnstileHandler handler =
-	    new TurnstileHandler();
-	TurnstileStepper stepper =
-	    new TurnstileStepper(handler);
+	Handler handler =
+	    new Handler();
+	Stepper stepper =
+	    new Stepper(handler);
 	stepper.step(Input.ticket);
 	stepper.step(Input.pass);
 	stepper.step(Input.ticket);

### 2013-11-28 languages/fsml/java/Handler.java

@@ -1,4 +1,16 @@
-// Reusable code
-public interface Handler<A> {
-    public void handle(A a);
+// Turnstile-specific, handwritten code 
+public class Handler implements HandlerBase<Action> {
+    public void handle(Action a) {
+        switch (a) {
+        case collect : 
+            System.out.println("collect");
+            break;
+        case alarm : 
+            System.out.println("alarm");
+            break;
+        case eject : 
+            System.out.println("eject");
+            break;
+        }
+    }
 }

### 2013-11-28 languages/fsml/java/Stepper.java

@@ -1,25 +1,15 @@
-import java.util.HashMap;
-
-// Reusable code
-public abstract class Stepper<S, I, A> {
-    protected S state;
-    protected Handler<A> handler;
-    private HashMap<S,HashMap<I,Pair<A,S>>> table =
-        new HashMap<S,HashMap<I,Pair<A,S>>>();
-    public final void add(S from, I i, A a, S to) {
-        if (!table.containsKey(from))
-            table.put(from, new HashMap<I,Pair<A,S>>());
-        HashMap<I,Pair<A,S>> subtable = table.get(from);
-        Pair<A,S> pair = new Pair<A,S>(a, to);
-        subtable.put(i, pair);
-    }
-    public final void step(I i) {
-        HashMap<I,Pair<A,S>> subtable = table.get(state);
-        Pair<A,S> pair = subtable.get(i);
-        S from = state;
-        S to = pair.y;
-        System.out.println("from: "+from+", input: "+i+", to: "+to);
-        if (pair.x!=null) handler.handle(pair.x);
-        state = to;
-    }
+// Generated code
+public class Stepper extends StepperBase<State, Input, Action> {
+  public Stepper(HandlerBase<Action> handler) {
+    this.handler = handler;
+    state = State.locked;
+    add(State.locked, Input.ticket, Action.collect, State.unlocked);
+    add(State.locked, Input.pass, Action.alarm, State.exception);
+    add(State.unlocked, Input.ticket, Action.eject, State.unlocked);
+    add(State.unlocked, Input.pass, null, State.locked);
+    add(State.exception, Input.ticket, Action.eject, State.exception);
+    add(State.exception, Input.pass, null, State.exception);
+    add(State.exception, Input.mute, null, State.exception);
+    add(State.exception, Input.release, null, State.locked);
+  }
 }

### 2013-11-28 languages/fsml/java/TurnstileHandler.java

@@ -1,16 +1,0 @@
-// Turnstile-specific, handwritten code 
-public class TurnstileHandler implements Handler<Action> {
-    public void handle(Action a) {
-        switch (a) {
-        case collect : 
-            System.out.println("collect");
-            break;
-        case alarm : 
-            System.out.println("alarm");
-            break;
-        case eject : 
-            System.out.println("eject");
-            break;
-        }
-    }
-}

### 2013-11-28 languages/fsml/java/TurnstileStepper.java

@@ -1,15 +1,0 @@
-// Generated code
-public class TurnstileStepper extends Stepper<State, Input, Action> {
-  public TurnstileStepper(Handler<Action> handler) {
-    this.handler = handler;
-    state = State.locked;
-    add(State.locked, Input.ticket, Action.collect, State.unlocked);
-    add(State.locked, Input.pass, Action.alarm, State.exception);
-    add(State.unlocked, Input.ticket, Action.eject, State.unlocked);
-    add(State.unlocked, Input.pass, null, State.locked);
-    add(State.exception, Input.ticket, Action.eject, State.exception);
-    add(State.exception, Input.pass, null, State.exception);
-    add(State.exception, Input.mute, null, State.exception);
-    add(State.exception, Input.release, null, State.locked);
-  }
-}

### 2013-12-25 docs/fsml/paper.tex
Enabled negative test cases and added force semantics to megamodel execution. Some code clean-up needed soon, but this is a pretty useful state of affairs now.

@@
 
-0.00002 as of 25 November 2013.
+0.00003 as of 25 December 2013.
 
@@
 possibly leveraging diverse technologies and techniques.
+
+Various technical details are separated out into the appendix.
 
@@
 summary. Constraint \emph{fsmSingleInitial} is meant to ensure that
-there is only a single state declaration for an initial
-state. Constraint \emph{fsmDistinctIds} is meant to ensure that the
-state ids of the state declarations are distinct. Constraint
-\emph{fsmResolvable} is meant to ensure that all referenced target
-states are declared. Constraint \emph{fsmDeterministic} is meant
-to ensure that each given input uniquely defines the target state to
-be transitioned to, if any. Constraint \emph{fsmReachable} is meant to
-ensure that all declared states are reachable from the
-initial state. In \autoref{languages/fsml/ok.pro}, all constraint
-predicates are surround by an application of the meta-predicate
-\emph{require/1}, which is simply there for better error
-reporting. That is, if the argument of \emph{require/1} fails, then
-this is reported immediately, as opposed to simply propagating failure
-silently.
+there is exactly one initial state declaration. Constraint
+\emph{fsmDistinctIds} is meant to ensure that the state ids of the
+state declarations are distinct. Constraint \emph{fsmResolvable} is
+meant to ensure that all referenced target states are
+declared. Constraint \emph{fsmDeterministic} is meant to ensure that
+each given input uniquely defines the target state to be transitioned
+to, if any. Constraint \emph{fsmReachable} is meant to ensure that all
+declared states are reachable from the initial state. In
+\autoref{languages/fsml/ok.pro}, all constraint predicates are
+surround by an application of the meta-predicate \emph{require/1},
+which is simply there for better error reporting. That is, if the
+argument of \emph{require/1} fails, then this is reported immediately,
+as opposed to simply propagating failure silently.
 
@@
 \codefigure{%
+A negative test case: \autoref{languages/fsml/sample.fsml} with a second initial state}{%
+languages/fsml/tests/initialNotOk.fsml}{%
+fsml}
+
+\codefigure{%
 The state ids of the state declarations are distinct}{%
@@
 prolog}
+
+\codefigure{%
+A negative test case: \autoref{languages/fsml/sample.fsml} with a
+double declaration of state `locked'}{%
+languages/fsml/tests/idsNotOk.fsml}{%
+fsml}
 
@@
 \codefigure{%
+A negative test case: \autoref{languages/fsml/sample.fsml} with a
+unresolvable state reference `unlockkked'}{%
+languages/fsml/tests/resolutionNotOk.fsml}{%
+fsml}
+
+\codefigure{%
 Each given input uniquely defines the target state to be transitioned to, if any}{%
@@
 prolog}
+
+\codefigure{%
+A negative test case: \autoref{languages/fsml/sample.fsml} with a
+nondeterministic transition; see the double occurrence of `ticket' in
+state `locked'}{%
+languages/fsml/tests/determinismNotOk.fsml}{%
+fsml}
 
@@
 \codefigure{%
+A negative test case: \autoref{languages/fsml/sample.fsml} with an
+unreachable state `unlocked'}{%
+languages/fsml/tests/reachabilityNotOk.fsml}{%
+fsml}
+
+\codefigure{%
 Fixed point computation needed in \autoref{languages/fsml/ok/reachable.pro}}{%
@@
 All constraints are specified in
-\autoref{languages/fsml/ok/initial.pro}--\autoref{languages/fsml/ok/closure.pro}. The
+\autoref{languages/fsml/ok/initial.pro}--\autoref{languages/fsml/ok/closure.pro}.
+We also show negative test cases that hence illustrate violation of
+the constraints; these cases are derived by minimalistic mutation of
+the sample FSM of \autoref{languages/fsml/sample.fsml}. The
 constraints are relatively simple to express using Prolog's
@@
 refactoring.
+
+\codefigure{%
+A possible input sequence for FSM of
+\autoref{languages/fsml/sample.fsml}}{%
+languages/fsml/sample.input}{%
+prolog}
+
+\codefigure{%
+The output for 
+\autoref{languages/fsml/sample.input}}{%
+languages/fsml/sample.output}{%
+prolog}
+
+Let us set up a test case for the FSM of
+\autoref{languages/fsml/sample.fsml}. That is, a valid input sequence
+for the FSM is shown in \autoref{languages/fsml/sample.input} and the
+corresponding output in terms of actions and state changes is shown in
+\autoref{languages/fsml/sample.input}. We shall define a semantics
+that indeed transforms the input into the output.
 
@@
 Generic interface for action handers}{%
-languages/fsml/java/Handler.java}{%
+languages/fsml/java/HandlerBase.java}{%
 java}
@@
 A trivial interpretation of the turnstile actions}{%
-languages/fsml/java/TurnstileHandler.java}{%
+languages/fsml/java/Handler.java}{%
 java}
 
-\autoref{languages/fsml/java/Handler.java} shows a generic, i.e.,
+\autoref{languages/fsml/java/HandlerBase.java} shows a generic, i.e.,
 FSM-independent interface for defining handlers for FSM
-actions. \autoref{languages/fsml/java/TurnstileHandler.java} shows a
+actions. \autoref{languages/fsml/java/Handler.java} shows a
 trivial implementation of said interface for the turnstile
@@
 Generic class for FSM execution}{%
-languages/fsml/java/Stepper.java}{%
+languages/fsml/java/StepperBase.java}{%
 java}
 
-\autoref{languages/fsml/java/Stepper.java} shows essentially the
+\autoref{languages/fsml/java/StepperBase.java} shows essentially the
 runtime for FSMs. This is generic code which is appropriately
@@
 keep track of and make use of a handler for the involved actions. The
-class is called \emph{Stepper} because its key method is the one for
+class is called \emph{StepperBase} because its key method is the one for
 performing a step (a transition) based on an actual input item.
@@
 Configuration of a stepper for the turnstile example}{%
-languages/fsml/java/TurnstileStepper.java}{%
+languages/fsml/java/Stepper.java}{%
 java}
 
-\autoref{languages/fsml/java/TurnstileStepper.java} shows the
-generated code for the turnstile-specific \emph{Stepper} subclass. The
-class instantiates the generic parameters of the \emph{Stepper} class
+\autoref{languages/fsml/java/Stepper.java} shows the generated code
+for the turnstile-specific subclass of the \emph{StepperBase}
+class. The generic parameters of \emph{StepperBase} are instantiated
 to the turnstile-specific enum types for states, inputs, and actions,
@@
 \codefigure{%
-Code generation for stepper subclass}{%
+Code generation for stepper class}{%
 languages/fsml/to-java/stepper.pro}{%
@@
 \autoref{languages/fsml/to-java/stepper.pro} shows the code generation
-predicate for stepper subclasses. The predicate refers to more
-elements of the Java language: several expression forms (e.g.,
-`null'), assignment statements, method calls, and class
-declarations. In a final step, the constructed abstract syntactical
-representation is passed to the pretty printer for Java.
+predicate for steppers. The predicate refers to more elements of the
+Java language: several expression forms (e.g., `null'), assignment
+statements, method calls, and class declarations. In a final step, the
+constructed abstract syntactical representation is passed to the
+pretty printer for Java.
 
@@
 are directed; nodes are of a certain shape (an ellipse here) and style
-(regular or
-filled). \autoref{languages/fsml/sample.dgl} represents
+(regular or filled). \autoref{languages/fsml/sample.dgl} represents
 represents the graph in the abstract (Prolog-based) syntax of a simple
@@
 
+\pagebreak
+
+\codefigure{%
+A negative test case: \autoref{languages/fsml/sample.fsml} with a syntax error; see `innnitial'}{%
+languages/fsml/tests/parserError.fsml}{%
+fsml}
+
+\codefigure{%
+A negative test case: an input for
+\autoref{languages/fsml/sample.fsml} which contains an invalid input symbol}{%
+languages/fsml/tests/illegalSymbol.input}{%
+prolog}
+
+\codefigure{%
+A negative test case: an input for
+\autoref{languages/fsml/sample.fsml} which contains an infeasible
+input symbol in the pool position---the initial state `locked' does
+not accept the `mute' input}{%
+languages/fsml/tests/infeasibleSymbol.input}{%
+prolog}
+
+\section{Additional test cases for the specification}
+\label{A:testing}
+
+Let us add a few more negative test cases in addition to the
+illustrations of constraint violation of \S\ref{S:ok}.  A simple
+parsing error is caused by the input of
+\autoref{languages/fsml/tests/parserError.fsml}.
+\autoref{languages/fsml/tests/illegalSymbol.input} and
+\autoref{languages/fsml/tests/infeasibleSymbol.input} demonstrate two
+major scenarios for failure of `simulation' according to
+\S\ref{S:semantics}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \end{document}

### 2014-10-01 languages/yafpl/sample.hs
More megamodels

@@ -1,0 +1,1 @@
+sample.yafpl

### 2014-10-02 docs/fsml/paper.tex
Slightly updated FSML spec paper

@@
 
-0.00003 as of 25 December 2013.
+0.00004 as of 2 October 2014.
 
@@
 \vspace{-77\in}
-\includegraphics[width=\textwidth]{../../languages/fsml/sample.pdf}
+\includegraphics[width=\textwidth]{../../languages/fsml/dot/sample.pdf}
 \vspace{-150\in}
@@
 The FSM of \autoref{F:turnstile} in the abstract syntax of the DGL language}{%
-languages/fsml/sample.dgl}{%
+languages/fsml/dot/sample.term}{%
 prolog}
@@
 The FSM of \autoref{F:turnstile} in the dot language}{%
-languages/fsml/sample.dot}{%
+languages/fsml/dot/sample.dgl}{%
 dot}
 
-The DGL processor translates the graph into a specification according
-to the \emph{dot} language of
+The DGL processor translates the graph into the concrete syntax
+of the \emph{dot} language of
 \emph{GraphViz}\footnote{\url{http://www.graphviz.org/}}; see
-\autoref{languages/fsml/sample.dot}. The GraphViz tool
+\autoref{languages/fsml/sample.dot}. (That is, DGL can be viewed as a 
+subset of the DOT language.) The GraphViz tool
 directly renders the dot specification as shown \autoref{F:turnstile}.

### 2014-12-29 docs/fsml/paper.tex
Working on the syntax chapter of the slbook

@@
 
-\title{Another DSL primer}
+\title{Model of a DSL for finite state machines}
 
@@
   development are parts of a broader effort on language modeling and
-  software language engineering---joint work with \emph{Anya Helene
-    Bagge}, University of Bergen. Helpful interaction with and
-  feedback by Andrei Varanovich, University of Koblenz-Landau, is also
-  gratefully acknowledged.}
+  software language engineering. Collaboration with and feedback by
+  Anya Helene Bagge (University of Bergen) and Andrei Varanovich
+  (University of Koblenz-Landau) are gratefully acknowledged.}
 
@@
 
-0.00004 as of 2 October 2014.
+0.1 as of 11 November 2014.
 
@@
 \noindent
-\textbf{Repository location of document}:
+\textbf{Website of this document}:
 
-\url{https://github.com/slebok/slepro/tree/master/docs/fsml}.
+\smallskip
+
+\url{http://softlang.uni-koblenz.de/fsml/}
+
+\smallskip
+
+\noindent
+The page also links to supplementary repository locations. In
+particular, the model (code) of this document as well as alternative
+implementations of FSML are linked. The FSML model is maintained as
+part of the \slepro{} project: \url{https://github.com/slebok/slepro}
 \end{abstract}
@@
 Various technical details are separated out into the appendix.
-
-The FSML development is part of the \slepro{} project. All software
-artifacts of this document and all support infrastructure is available
-through the repository of the project:
-
-\begin{center}
-\url{https://github.com/slebok/slepro}
-\end{center}
 

### 2014-12-29 languages/bnl/esl/java/Number.java

@@ -1,0 +1,6 @@
+import java.util.List;
+
+public class Number {
+    public List<Bit> bits;
+    public List<Bit> rest;
+}

### 2014-12-29 languages/bnl/java/Rational.java

@@ -1,0 +1,3 @@
+public class Rational extends Rest {
+    Bits bits
+}

### 2014-12-30 languages/bfpl/functional/efpl/Main.hs
slepro is retiring; we still need to find a new home for the fsml doc, though.

@@ -1,1 +1,0 @@
-Main.efpl

### 2014-12-30 languages/bfpl/functional/efpl/Sample.hs

@@ -1,1 +1,0 @@
-Sample.efpl

### 2014-12-30 languages/bfpl/functional/efpl/Semantics.hs

@@ -1,1 +1,0 @@
-Semantics.efpl

### 2014-12-30 languages/bfpl/functional/efpl/Stdlib.hs

@@ -1,1 +1,0 @@
-Stdlib.efpl

### 2014-12-30 languages/bfpl/functional/efpl/Syntax.hs

@@ -1,1 +1,0 @@
-Syntax.efpl

### 2014-12-30 languages/bfpl/functional/efpl/Types.hs

@@ -1,1 +1,0 @@
-Types.efpl

### 2014-12-30 languages/bfpl/sample.hs

@@ -1,1 +1,0 @@
-sample.bfpl

### 2014-12-30 languages/bnl/esl/java/Number.java

@@ -1,6 +1,0 @@
-import java.util.List;
-
-public class Number {
-    public List<Bit> bits;
-    public List<Bit> rest;
-}

### 2014-12-30 languages/bnl/java/Rational.java

@@ -1,3 +1,0 @@
-public class Rational extends Rest {
-    Bits bits
-}

### 2014-12-30 languages/fsml/java/Action.java

@@ -1,2 +1,0 @@
-// Generated code
-public enum Action {collect, alarm, eject}

### 2014-12-30 languages/fsml/java/Demo.java

@@ -1,23 +1,0 @@
-// Turnstile-specific, handwritten code
-public class Demo {
-    public static void main(String [] args) {
-	Handler handler =
-	    new Handler();
-	Stepper stepper =
-	    new Stepper(handler);
-	stepper.step(Input.ticket);
-	stepper.step(Input.pass);
-	stepper.step(Input.ticket);
-	stepper.step(Input.pass);
-	stepper.step(Input.ticket);
-	stepper.step(Input.ticket);
-	stepper.step(Input.pass);
-	stepper.step(Input.pass);
-	stepper.step(Input.ticket);
-	stepper.step(Input.pass);
-	stepper.step(Input.mute);
-	stepper.step(Input.release);
-	stepper.step(Input.ticket);
-	stepper.step(Input.pass);
-    }
-}

### 2014-12-30 languages/fsml/java/Handler.java

@@ -1,16 +1,0 @@
-// Turnstile-specific, handwritten code 
-public class Handler implements HandlerBase<Action> {
-    public void handle(Action a) {
-        switch (a) {
-        case collect : 
-            System.out.println("collect");
-            break;
-        case alarm : 
-            System.out.println("alarm");
-            break;
-        case eject : 
-            System.out.println("eject");
-            break;
-        }
-    }
-}

### 2014-12-30 languages/fsml/java/HandlerBase.java

@@ -1,4 +1,0 @@
-// Reusable code
-public interface HandlerBase<A> {
-    public void handle(A a);
-}

### 2014-12-30 languages/fsml/java/Input.java

@@ -1,2 +1,0 @@
-// Generated code
-public enum Input {ticket, pass, mute, release}

### 2014-12-30 languages/fsml/java/Pair.java

@@ -1,9 +1,0 @@
-// Reusable code
-public class Pair<X,Y> {
-    public X x;
-    public Y y;
-    public Pair(X x, Y y) {
-	this.x = x;
-	this.y = y;
-    }
-}

### 2014-12-30 languages/fsml/java/State.java

@@ -1,2 +1,0 @@
-// Generated code
-public enum State {locked, unlocked, exception}

### 2014-12-30 languages/fsml/java/Stepper.java

@@ -1,15 +1,0 @@
-// Generated code
-public class Stepper extends StepperBase<State, Input, Action> {
-  public Stepper(HandlerBase<Action> handler) {
-    this.handler = handler;
-    state = State.locked;
-    add(State.locked, Input.ticket, Action.collect, State.unlocked);
-    add(State.locked, Input.pass, Action.alarm, State.exception);
-    add(State.unlocked, Input.ticket, Action.eject, State.unlocked);
-    add(State.unlocked, Input.pass, null, State.locked);
-    add(State.exception, Input.ticket, Action.eject, State.exception);
-    add(State.exception, Input.pass, null, State.exception);
-    add(State.exception, Input.mute, null, State.exception);
-    add(State.exception, Input.release, null, State.locked);
-  }
-}

### 2014-12-30 languages/fsml/java/StepperBase.java

@@ -1,25 +1,0 @@
-import java.util.HashMap;
-
-// Reusable code
-public abstract class StepperBase<S, I, A> {
-    protected S state;
-    protected HandlerBase<A> handler;
-    private HashMap<S,HashMap<I,Pair<A,S>>> table =
-        new HashMap<S,HashMap<I,Pair<A,S>>>();
-    public final void add(S from, I i, A a, S to) {
-        if (!table.containsKey(from))
-            table.put(from, new HashMap<I,Pair<A,S>>());
-        HashMap<I,Pair<A,S>> subtable = table.get(from);
-        Pair<A,S> pair = new Pair<A,S>(a, to);
-        subtable.put(i, pair);
-    }
-    public final void step(I i) {
-        HashMap<I,Pair<A,S>> subtable = table.get(state);
-        Pair<A,S> pair = subtable.get(i);
-        S from = state;
-        S to = pair.y;
-        System.out.println("from: "+from+", input: "+i+", to: "+to);
-        if (pair.x!=null) handler.handle(pair.x);
-        state = to;
-    }
-}

### 2014-12-30 languages/java/sample.java

@@ -1,2 +1,0 @@
-// Generated code
-public enum foo {v1, v2, v3}
